ProteInfer, deep neural networks for protein functional inference

### 2023-01-16 latex/draft3.tex
Updates
updates

%%%TOPSTUFF%%%
\title{ProteInfer: deep networks for protein functional inference}
\shorttitle{ProteInfer}
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[*,1,\Letter]{Theo Sanderson}
\author[*,2]{Maxwell L. Bileschi}
\author[2]{David Belanger}
\author[2]{Lucy J. Colwell}

@@

\begin{abstract}
  Predicting the function of a protein from its amino acid sequence is a long-standing challenge in bioinformatics. Traditional approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. Here we instead employ deep convolutional neural networks to directly predict a variety of protein functions -- EC numbers and GO terms -- directly from an unaligned amino acid sequence. This approach provides precise predictions which complement alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces, which we demonstrate with an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. To read the interactive version of this paper, please visit \url{https://google-research.github.io/proteinfer/}
\end {abstract}

@@

Of particular relevance to the present work is Bileschi et al. (2019) \citep{bileschi}, where it is shown that models with residual layers \citep{resnet} of dilated convolutions \citep{dilatedconvolutions} can precisely and efficiently categorise protein domains.  Dohan (2021) \cite{dohan2021kdd} provides additional accuracy improvements using uncurated data. However, these models cannot infer functional annotations for full-length protein sequences, since they are trained on pre-segmented domains and can only predict a single label. The full-sequence task is of primary importance to biological practitioners.

To address this challenge we employ deep dilated convolutional networks to learn the mapping between full-length protein sequences and functional annotations. The resulting ProteInfer models take amino acid sequences as input and are trained on the well-curated portion of the protein universe annotated by Swiss-Prot \cite{uniprotlatest}. We find that: 1) ProteInfer models reproduce curator decisions for a  variety of functional properties across sequences distant from the training data, 2) attribution analysis shows that the predictions are driven by relevant regions of each protein sequence, and 3) ProteInfer models create a generalised mapping between sequence space and the space of protein functions, which is useful for tasks other than those for which the models were trained. We provide trained ProteInfer networks that enable other researchers to reproduce the analysis presented and to explore embeddings of their proteins of interest, via both a command line tool\footnote{\url{https://github.com/google-research/proteinfer}}, and also via an in-browser JavaScript implementation that demonstrates the computational efficiency of deep-learning approaches.
\section*{Methods}
\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{Figures/figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}
%%%MODEL%%%
In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities. 
We select all labels with predicted probability above a given confidence threshold, and varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

@@

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot \footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}}. Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation. 
\section*{Results}
\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/ec_performance_by_top_level_class.png}%
  \caption{\textbf{ProteInfer performance for predictions within all 7 top-level enzyme groups.}}
  \label{fig:toplevel}
\end{figure}

@@

  \begin{figure*}
    \centering
      \includegraphics[width=\textwidth]{Figures/umap2.pdf}%
      \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
      \label{fig:umap}
    \end{figure*}

Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963) (\Cref{fig:toplevel}A%htmlver: 
). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

BLASTp is arguably a practitioner's default choice for functional annotation, so we implemented an alignment-based baseline in which BLASTp is used to identify the closest sequence to a query sequence in the train set. Labels are then imputed for the query sequence by transferring those labels that apply to the annotated match from the train set.

We produced a precision-recall curve by using the bit score of the closest sequence as a measure of confidence, varying the cutoff above which we retain the imputed labels \cite{zhou2019cafa, eddy2011accelerated}. We also considered an ensemble of neural networks \cite{bileschi}, where the average of the ensemble elements' predicted probabilities is used as a confidence score. %
%startleaveout
(See \Cref{fig:with_ensemble}, \Cref{fig:without_ensemble}.)
%stopleaveout

%%%METHOD%%%
We found that BLASTp was able to achieve higher recall values than ProteInfer for lower precision values, while ProteInfer was able to provide greater precision than BLASTp at lower recall values. We wondered whether a combination of ProteInfer and BLASTp could synergize the best properties of both approaches. We found that even the simple ensembling strategy of rescaling the BLAST bit score by the averages of the ensembled CNNs' predicted probabilities gave a $\texttt{F}_{max}$ score (0.991, 95\% confidence interval [CI]: 0.990--0.992 ) that exceeded that of BLAST (0.984, 95\% CI: 0.983--0.985) or the ensembled CNN (0.981, 95\% CI: 0.980--0.982) alone. 
On the clustered train-test split based on UniRef50 (see \textit{clustered} in \Cref{fig:toplevel}B%htmlver:<a href="#fig5" target="_self">Fig. 5</a>
), we see a performance drop in all methods: this is expected, as remote homology tasks are designed to challenge methods to generalize farther in sequence space. The $\texttt{F}_{max}$ score of a single neural network fell to 0.914 (95\% CI: 0.913--0.915, precision: 0.959 recall: 0.875), substantially lower than BLAST (0.950, 95\% CI: 0.950--0.951), though again an ensemble of both BLAST and ProteInfer outperformed both (0.979, 95\% CI: 0.979--0.980). We find that neural network methods learn different information about proteins than alignment-based methods, and a combination of the two further improves remote homology detection.

@@

%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} This figure shows EC-trained ProteInfer embeddings for all non-enzymatic sequences in the test set, projected using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, Q8RUD6 is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). Q8RUD6 can however act as a sulfertransferase \cite{pmid17408957}.

@@

There is arguably room for improvement in this regard from classical approaches. For example, the online interface to InterProScan can take ~147 seconds to process a 1500 amino acid sequence\footnote{Protein Q77Z83.}\footnote{It should be noted that the individual databases that make up InterProScan may return matches faster, with the online interface to Pfam taking 14--20 seconds for a 1500 amino acid sequence.}, while running the tool may make the search faster, doing so requires downloading a 9 GB file, with an additional 14 GB for the full set of signatures, which when installed exceeds 51 GB. Meanwhile, conducting a BLAST search against Swiss-Prot takes 34 seconds for a 1500 amino acid sequence \footnote{A target database with decreased redundancy could be built to reduce this search time, and other optimizations of BLAST have been developed.}.

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (\Cref{sup:js_timing}%htmlver:see <a href="./latex/supplement.pdf">supplement</a>
).
% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}[t]
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}

To assess our model's performance using an additional experimentally-validated source of ground truth, we focused our attention on a large set of bacterial genes for which functions have recently been identified in a high-throughput experimental genetic study \citep{phenotypes}. In particular, this study listed newly identified EC numbers for 171 proteins, representing cases when there was previously either misannotation or inconsistent annotation in the SEED or KEGG databases. Therefore, this set of genes may be enriched for proteins whose functions are difficult to assess computationally.
We examined how well our network was able to make predictions for this experimental dataset at each level of the EC hierarchy (\Cref{fig:levels}%htmlver:<a href="#fig10" target="_self">Fig. 11</a>
), using as a decision threshold the value that optimised F1 identified during tuning. The network had high accuracy for identification of broad enzyme class, with 90\% accuracy at the top level of the EC hierarchy. To compute accuracy, we examined the subset of these 171 proteins for which there was a single enzymatic annotation from \cite{phenotypes}, giving us predictions for 119 enzymes. At the second level of the hierarchy, accuracy was 90\% and the network declined to make a prediction for 12\% of classes. Even at the third level, accuracy was 86\% with the network making a prediction in 77\% of cases. At the finest level of classification, the proportion of examples for which a prediction was made fell to 28\%, with 42\% of these predictions correct.

@@

\subsection*{A model predicting the entire gene ontology}
Given the high accuracy that our deep learning model was able to achieve on the more than five thousand enzymatic labels in Swiss-Prot, we asked whether our networks could learn to predict protein properties using an even larger vocabulary of labels, using a similar test-train setup. Gene Ontology \citep{ashburner2000gene, gene2019gene, carbon2009amigo} (GO) terms describe important protein functional properties, with 28,079 such terms in Swiss-Prot that cover the molecular functions of proteins (e.g. DNA-binding, amylase activity), the biological processes they are involved in (e.g. DNA replication, meiosis), and the cellular components to which they localise (e.g. mitochondrion, cytosol). These terms are arranged in a complex directed acyclic graph, with some nodes having as many as 12 ancestral nodes.

@@

\onecolumn
\newpage
%stopleaveout

>>>>

%%%TOPSTUFF%%%
\title{ProteInfer, deep neural networks for protein functional inference}
\shorttitle{ProteInfer}
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[*,1,\Letter]{Theo Sanderson}
\author[*,2]{Maxwell L. Bileschi}
\author[2]{David Belanger}
\author[2]{Lucy J. Colwell}

@@

\begin{abstract}
  Predicting the function of a protein from its amino acid sequence is a long-standing challenge in bioinformatics. Traditional approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. Here we introduce ProteInfer, which instead employs deep convolutional neural networks to directly predict a variety of protein functions -- EC numbers and GO terms -- directly from an unaligned amino acid sequence. This approach provides precise predictions which complement alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces, which we demonstrate with an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. To read the interactive version of this paper, please visit \url{https://google-research.github.io/proteinfer/}.
\end {abstract}

@@

Of particular relevance to the present work is Bileschi et al. (2022) \citep{bileschi}, where it is shown that models with residual layers \citep{resnet} of dilated convolutions \citep{dilatedconvolutions} can precisely and efficiently categorise protein domains.  Dohan (2021) \cite{dohan2021kdd} provides additional accuracy improvements using uncurated data. However, these models cannot infer functional annotations for full-length protein sequences, since they are trained on pre-segmented domains and can only predict a single label. The full-sequence task is of primary importance to biological practitioners.

To address this challenge we employ deep dilated convolutional networks to learn the mapping between full-length protein sequences and functional annotations. The resulting ProteInfer models take amino acid sequences as input and are trained on the well-curated portion of the protein universe annotated by Swiss-Prot \cite{uniprotlatest}. We find that: 1) ProteInfer models reproduce curator decisions for a  variety of functional properties across sequences distant from the training data, 2) attribution analysis shows that the predictions are driven by relevant regions of each protein sequence, and 3) ProteInfer models create a generalised mapping between sequence space and the space of protein functions, which is useful for tasks other than those for which the models were trained. We provide trained ProteInfer networks that enable other researchers to reproduce the analysis presented and to explore embeddings of their proteins of interest, via both a command line tool\footnote{\url{https://github.com/google-research/proteinfer}}, and also via an in-browser JavaScript implementation that demonstrates the computational efficiency of deep-learning approaches.
\section*{Results}

\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{Figures/figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}
%%%MODEL%%%
In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. In contrast, many previous approaches operate on fixed sequence lengths: these techniques are unable to make predictions for proteins larger than this sequence length, and use unnecessary resources when employed on smaller proteins. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities. 
We select all labels with predicted probability above a given confidence threshold, and varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

@@

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot.\footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}} Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation.

\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/fig3}%
  \caption{\textbf{ProteInfer performance (A) for all 7 top-level enzyme groups from a single CNN model (B) compared between methods: a single ProteInfer CNN, an ensemble of Proteinfer CNNs, a BLAST-based baseline, and an ensemble of BLAST predictions combined with ProteInfer CNNs }}
  \label{fig:toplevel}
\end{figure}

@@

  \begin{figure*}
    \centering
      \includegraphics[width=\textwidth]{Figures/umap3.pdf}%
      \caption{\textbf{Embedding reflects enzyme functional hierarchy}. UMAP projection of embeddings for the subset of test set sequences which have only one leaf-level EC classification. Points are colour-coded at successive levels of the EC hierarchy in each panel. (A) colours denote top level EC groups, (B) colours denote second level EC groups within EC2.*, (C) colours denote third level EC groups within EC:2.7.*, (D) colours depict terminal EC groups within EC:2.7.4.* }
      \label{fig:umap}
    \end{figure*}

Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

We implemented an alignment-based baseline in which BLASTp is used to identify the closest sequence to a query sequence in the train set. Labels are then imputed for the query sequence by transferring those labels that apply to the annotated match from the train set. We produced a precision-recall curve by using the bit score of the closest sequence as a measure of confidence, varying the cutoff above which we retain the imputed labels \cite{zhou2019cafa, eddy2011accelerated}. We also considered an ensemble of neural networks \cite{bileschi}, where the average of the ensemble elements' predicted probabilities is used as a confidence score, and a naive control where the number of proteins annotated with a specific term in the training set \citep{radivojac2013large}. %
%startleaveout
(See \Cref{fig:toplevel}B, \Cref{fig:with_ensemble}, \Cref{fig:without_ensemble}.)
%stopleaveout

%%%METHOD%%%
We found that BLASTp was able to achieve higher recall values than ProteInfer for lower precision values, while ProteInfer was able to provide greater precision than BLASTp at lower recall values. The high recall of BLAST is likely to reflect the fact that it has access to the entirety of the training set, rather than having to compress it into a limited set of neural network weights. In contrast, the lack of precision in BLAST could relate to reshuffling of sequences during evolution, which would allow a given protein to show high similarity to a trainining sequence in a particular subregion, despite lacking the core region required for that training sequence's function.  We wondered whether a combination of ProteInfer and BLASTp could synergize the best properties of both approaches. We found that even the simple ensembling strategy of rescaling the BLAST bit score by the averages of the ensembled CNNs' predicted probabilities gave a $\texttt{F}_{max}$ score (0.991, 95\% confidence interval [CI]: 0.990--0.992 ) that exceeded that of BLAST (0.984, 95\% CI: 0.983--0.985) or the ensembled CNN (0.981, 95\% CI: 0.980--0.982) alone (see Supplement for more details on this method). 
On the clustered train-test split based on UniRef50 (see \textit{clustered} in \Cref{fig:toplevel}B%htmlver:<a href="#fig5" target="_self">Fig. 5</a>
), we see a performance drop in all methods: this is expected, as remote homology tasks are designed to challenge methods to generalize farther in sequence space. The $\texttt{F}_{max}$ score of a single neural network fell to 0.914 (95\% CI: 0.913--0.915, precision: 0.959 recall: 0.875), substantially lower than BLAST (0.950, 95\% CI: 0.950--0.951), though again an ensemble of both BLAST and ProteInfer outperformed both (0.979, 95\% CI: 0.979--0.980). These patterns suggest that neural network methods learn different information about proteins to alignment-based methods, and so a combination of the two provides a synergistic result. All methods dramatically outperformed the naive frequency-based approach (\Cref{sup:fig:naive}%htmlver:see supplement
).

@@

%%%embedding_explorer%%%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/nucleotidebinding}
        \caption{Nucleotide binding}
        \label{fig:gull}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/ribosome}
        \caption{Structural constituent of ribosome}
        \label{fig:tiger}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.3\textwidth}
    \vskip 0pt
        \includegraphics[width=\textwidth]{Figures/membrane}
        \caption{Intrinsic component of membrane}
        \label{fig:mouse}
    \end{subfigure}
    \caption{\textbf{A neural network trained on enzyme function learns general protein properties, beyond enzymatic activity.} This figure shows EC-trained ProteInfer embeddings for all non-enzymatic sequences in the test set, projected using UMAP. To illustrate the structure contained in these embeddings we highlight genes based on GO labels (on which this network was never trained).}\label{fig:unsupervised_embedding}
\end{figure*}
The resulting representation captures the hierarchical nature of EC classification, with the largest clusters in embedding space corresponding to top level EC groupings (\Cref{fig:umap}A%htmlver:
). These clusters in turn are further divided into sub-regions on the basis of subsequent levels of the EC hierarchy (\Cref{fig:umap}B%%htmlver:
). Exceptions to this rule generally recapitulate biological properties. For instance, \emph{Q8RUD6} is annotated as Arsenate reductase (glutaredoxin) (EC:1.20.4.1) \cite{pmid25464340} was not placed with other oxidoreductases (EC:1.-.-.-) but rather with Sulfurtransferases (EC:2.8.1.-). \emph{Q8RUD6} can however act as a sulfertransferase \cite{pmid17408957}.

@@

There is arguably room for improvement in this regard from classical approaches. For example, the online interface to InterProScan can take ~147 seconds to process a 1500 amino acid sequence\footnote{Protein used was Q77Z83. It should be noted that the individual databases that make up InterProScan may return matches faster, with the online interface to Pfam taking 14--20 seconds for a 1500 amino acid sequence.}, while running the tool may make the search faster, doing so requires downloading a 9 GB file, with an additional 14 GB for the full set of signatures, which when installed exceeds 51 GB. Meanwhile, conducting a BLAST search against Swiss-Prot takes 34 seconds for a 1500 amino acid sequence \footnote{A target database with decreased redundancy could be built to reduce this search time, and other optimizations of BLAST have been developed.}.

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer.  We selected the hyperparameters for these lightweight models by performing a tuning study in which we filtered results by the size of the model's parameters and then selected the best performing models. This approach uses a single neural network, rather than an ensemble. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (see supplement).

% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}[t]
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}

Despite its curated nature, SwissProt contains many proteins annotated only on the basis of electronic tools. To assess our model's performance using an experimentally-validated source of ground truth, we focused our attention on a large set of bacterial genes for which functions have recently been identified in a high-throughput experimental genetic study \citep{phenotypes}. In particular, this study listed newly identified EC numbers for 171 proteins, representing cases when there was previously either misannotation or inconsistent annotation in the SEED or KEGG databases. Therefore, this set of genes may be enriched for proteins whose functions are difficult to assess computationally.
We examined how well our network was able to make predictions for this experimental dataset at each level of the EC hierarchy (\Cref{fig:levels}%htmlver:<a href="#fig10" target="_self">Fig. 11</a>
), using as a decision threshold the value that optimised F1 identified during tuning. The network had high accuracy for identification of broad enzyme class, with 90\% accuracy at the top level of the EC hierarchy. To compute accuracy, we examined the subset of these 171 proteins for which there was a single enzymatic annotation from \cite{phenotypes}, giving us predictions for 119 enzymes. At the second level of the hierarchy, accuracy was 90\% and the network declined to make a prediction for 12\% of classes. Even at the third level, accuracy was 86\% with the network making a prediction in 77\% of cases. At the finest level of classification, the proportion of examples for which a prediction was made fell to 28\%, with 42\% of these predictions correct.

@@

\subsection*{A model predicting the entire gene ontology}
Given the high accuracy that our deep learning model was able to achieve on the more than five thousand enzymatic labels in Swiss-Prot, we asked whether our networks could learn to predict protein properties using an even larger vocabulary of labels, using a similar test-train setup. Gene Ontology \citep{ashburner2000gene, gene2019gene, carbon2009amigo} (GO) terms describe important protein functional properties, with 32,109 (\Cref{sup:table:vocab_sizes}%htmlver:see supplement
) such terms in Swiss-Prot that cover the molecular functions of proteins (e.g. DNA-binding, amylase activity), the biological processes they are involved in (e.g. DNA replication, meiosis), and the cellular components to which they localise (e.g. mitochondrion, cytosol). These terms are arranged in a complex directed acyclic graph, with some nodes having as many as 12 ancestral nodes.

@@

\onecolumn
\newpage

\section*{Supplement}
\beginsupplement
